---
title: "NGS data analysis of the Bac71-23 libraries"
author: "Philipp Koch"
date: "06.06.2021"
output:
  word_document: default
  pdf_document: default
subtitle: DESeq2, shrunken log fold changes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = FALSE)
```

## Load packages and data

```{r}
library(tidyverse)
library(DESeq2)
```

Load Supplementary Table 1 & 2, containing NGS read counts for each peptide-encoding DNA
```{r}
#for the random library
countData_random <- read.csv("../Supplementary Tables/Supplementary Table 1_NGS_random.csv") %>%
  column_to_rownames("ID") %>%
  select(-Sequence) %>%
  as.matrix()

#for the site saturation library
countData_sitesat <- read.csv("../Supplementary Tables/Supplementary Table 3_NGS_sitesat.csv") %>%
  column_to_rownames("ID") %>%
  select(-Sequence) %>%
  as.matrix()
```

## Pre-process

DESeq2 uses the first level of `timepoint` as the baseline for tests. Since we want to compare against $t_0$, we relevel this to be the first level of `timepoint` (also necessary for using `lfcShrink` later). We leave $t_4$ as the last level, since this is the condition of interest.

```{r}
#for the random library
timepoint_random <- parse_number(colnames(countData_random)) %>% factor() %>% relevel("0")
batch_random <- sub(".*\\d", "", colnames(countData_random))
colData_random <- data.frame(row.names = colnames(countData_random),
                      timepoint = timepoint_random,
                      batch = factor(batch_random))

#for the site saturation library
timepoint_sitesat <- parse_number(colnames(countData_sitesat)) %>% factor() %>% relevel("0")
batch_sitesat <- sub(".*\\d", "", colnames(countData_sitesat))
colData_sitesat <- data.frame(row.names = colnames(countData_sitesat),
                      timepoint = timepoint_sitesat,
                      batch = factor(batch_sitesat))
```

## Differential abundance analysis

DESeq2 uses the last term of the design formula for contrasts, which in our case should be `timepoint`.

```{r}
#for the random library
dds_random <- DESeqDataSetFromMatrix(countData_random, colData_random, ~ batch + timepoint)
dds_random <- dds_random[rowSums(counts(dds_random)) > 1,]

#for the site saturation library
dds_sitesat <- DESeqDataSetFromMatrix(countData_sitesat, colData_sitesat, ~ batch + timepoint)
dds_sitesat <- dds_sitesat[rowSums(counts(dds_sitesat)) > 1,]
```


Fit DESeq model. Data from all time points and replicates are used to estimate dispersion, but only the contrast $t_3$, corresponding to 4.5 h past induction, vs $t_0$, the time point for induction will be computed and tested in the end. Use `apeglm` to compute shrunken log2FoldChanges at time point 4 vs 0.

```{r}
#for the random library
dds.Wald_random <- DESeq(dds_random, parallel = TRUE)
res.Wald_random  <- results(dds.Wald_random , name = "timepoint_4_vs_0", altHypothesis = "less",
                    alpha = 0.05, parallel = TRUE)
lfcShrink_random  <- lfcShrink(dds.Wald_random , coef = "timepoint_4_vs_0", type = "apeglm",parallel = TRUE)
res.Wald_random$lfcShrink4 <- lfcShrink_random$log2FoldChange


#for the site saturation library
dds.Wald_sitesat <- DESeq(dds_sitesat, parallel = TRUE)
res.Wald_sitesat  <- results(dds.Wald_sitesat , name = "timepoint_4_vs_0", altHypothesis = "less",
                    alpha = 0.05, parallel = TRUE)
lfcShrink_sitesat  <- lfcShrink(dds.Wald_sitesat , coef = "timepoint_4_vs_0", type = "apeglm",parallel = TRUE)
res.Wald_sitesat$lfcShrink4 <- lfcShrink_sitesat$log2FoldChange
```

<!-- Plot a PCA of the data normalised using the regularised logarithmic transform `rlog()`. -->

```{r eval=FALSE, include=FALSE}
#for the random library
rld_random <- rlog(dds_random)
plotPCA(rld_random, intgroup = "timepoint") +
  theme_bw() +
  labs(title = "Principal component analysis of epPCR NGS count data",
       subtitle = "Counts normalised using the regularised log transform")

#for the site saturation library
rld_sitesat <- rlog(dds_sitesat)
plotPCA(rld_random, intgroup = "timepoint") +
  theme_bw() +
  labs(title = "Principal component analysis of site saturation NGS count data",
       subtitle = "Counts normalised using the regularised log transform")
```

## Save data

Rank list of peptides according to the shrunken log2FoldChange and save table as Supplementary Table 2 & 4
```{r}
#for the random library
results_random = as.data.frame(res.Wald_random) %>%
  rownames_to_column("ID") %>%
  left_join(read.csv("../Supplementary Tables/Supplementary Table 1_NGS_random.csv") %>% select(ID,Sequence) %>% mutate(ID = as.character(ID)),   by='ID') %>%
  select(ID,Sequence,lfcShrink4) %>%
  arrange(lfcShrink4)%>%
  mutate(Rank = c(1:nrow(.)))

#for the site saturation library
results_sitesat = as.data.frame(res.Wald_sitesat) %>%
  rownames_to_column("ID") %>%
  left_join(read.csv("../Supplementary Tables/Supplementary Table 3_NGS_sitesat.csv") %>% select(ID,Sequence) %>% mutate(ID = as.character(ID)),   by='ID') %>%
  select(ID,Sequence,lfcShrink4) %>%
  arrange(lfcShrink4)%>%
  mutate(Rank = c(1:nrow(.)))
                
                
#saveRDS(results_random,file='../Supplementary Tables/Supplementary Table 2.rds')
#saveRDS(results_sitesat,file='../Supplementary Tables/Supplementary Table 4.rds')

```
